---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![R-CMD-check](https://github.com/dmurdoch/tables/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dmurdoch/tables/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tables

The goal of tables is to compute and display complex tables of summary statistics.

Output may be in LaTeX, HTML, plain text, or an R
matrix for further processing.

## Installation

You can install the release version of `orientlib` using

``` r
install.packages("tables")
```

You can install the development version of tables from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dmurdoch/tables")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(tables)

# In an R Markdown document, you don't want each table
# to output the HTML document header, so turn 
# off that option:

table_options(htmloptions(head=FALSE))

X <- rnorm(125, sd=100)
Group <- factor(sample(letters[1:5], 125, rep=TRUE))
tab <- tabular( Group ~ 
                  (N=1) +
                  Format(digits=2)*X*
                    ((Mean=mean) +
                     Heading("Std Dev")*sd) 
              )

# To print in plain text:
tab

# To format in HTML:
toHTML(tab)

# To generate LaTeX code:
cat(toLatex(tab)$text)
```
