---
title: "HTML Tables"
author: "Duncan Murdoch"
date: "January 17, 2016"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{HTML Tables}
  \usepackage[utf8]{inputenc}
---

```{r echo = FALSE}
if (!requireNamespace("rmarkdown") || !rmarkdown::pandoc_available("1.12.3")) {
  warning("This vignette requires pandoc version 1.12.3; code will not run in older versions.")
  knitr::opts_chunk$set(eval = FALSE)
}
```
This short R Markdown document illustrates how
to use `tables` in HTML. 

## Initializing

First, you need to load `tables`.  Packages
that it uses print banner messages, so you'll 
usually want to set a chunk option of `message=FALSE`.
```{r results='hide', message=FALSE}
library(tables)
```

The table is now constructed in the usual way.  
```{r}
X <- rnorm(125, sd=100)
Group <- factor(sample(letters[1:5], 125, rep=TRUE))
tab <- tabular( Group ~ (N=1)+Format(digits=2)*X*((Mean=mean) + Heading("Std Dev")*sd) )
```
In an R Markdown document, you don't want each table
to output the HTML document header, so turn 
off those options:
```{r}
table_options(htmloptions(head=FALSE))
```

## Inserting a Table

There are two ways to insert a table into a knitr/rmarkdown document.
Since tables version 0.8.4, you can just print the object, and it
will be automatically formatted according to the current `table_options()`:
```{r}
tab
```

The previous behaviour  required an explicit call to the
`toHTML()` function, and knitr option 
`results = 'asis'` for the chunk.  To *require* the older method,
use `table_options(knit_print = FALSE)`:
```{r}
table_options(knit_print = FALSE)
tab        # This chunk uses the default results = 'markup'
```
```{r results = 'asis'}
toHTML(tab)  # This chunk uses results = 'asis'
```

In either case, some styling information needs to be included
in the document.  Ideally this should be in the header, but
it can be inserted anywhere by calling `writeCSS` with 
`results = 'asis'`:
```{r results='asis'}
writeCSS()
```

## Improving the Look

The default justification makes the columns of 
numbers look messy.  You can set the justification
to the right, but the headers look wrong:
```{r}
table_options(htmloptions(head = FALSE, justification = "r", knit_print = TRUE))
tab
```

The best look comes with the `pad = TRUE` option. 
This adds nonbreaking spaces around the numbers so that 
centering looks good.  It also changes the hyphens
to proper minus signs:
```{r}
table_options(htmloptions(head = FALSE, justification = "c", pad = TRUE))
tab
```
Unfortunately, if you cut this table and paste it
into a spreadsheet, the spaces and minus signs
probably won't be understood.  I don't know 
how to get everything we want :-(.

## Fine tuning

This document uses the default CSS from `table_options()$CSS`.  If you are
producing an `html_document`, it should be okay.
It does not look quite right in a
`slidy_presentation`, and is no good at all in
an `ioslides_presentation`.  Furthermore, you 
might not agree with my design choices.

In any of these cases, you should substitute your
own CSS.  You will need to modify the default 
one, and can use it as the `CSS` argument to 
`writeCSS()`, or set it as a new default in
`table_options()`.

Here is the default setting:
```{r comment=NA, echo=FALSE}
cat(table_options()$CSS)
```
Note that the `#ID` values will be replaced with
the `id` string given in `writeCSS()`, and `toHTML()` has a corresponding
argument to allow you to make changes for one specific table.
If you are using the newer form of display with
`table_options(knit_print = TRUE)`, you can get customized 
display for one table by also using `table_options(doCSS = TRUE)`.
For example,
```{r}
table_options(CSS =
"<style>
#ID .center { 
  text-align:center;
  background-color: aliceblue;
}
</style>", doCSS = TRUE)
tab
table_options(doCSS = FALSE)
tab
```

## kableExtra support

The kableExtra package contains a large number of functions
to customize the look of tables generated by the `knitr::kable()`
function.  These can also be made to work with tables from
this package, using the `toKable()` function.  For example,
```{r}
library(magrittr)
library(kableExtra)
toKable(tab, format="html") %>% 
  kable_styling("striped", position = "float_right", full_width = FALSE) %>%
  add_header_above(c("Row Label" = 1, "Statistics" = 3)) %>%
  column_spec(4, color = "red") %>%
  row_spec(1, color = "blue") %>%
  group_rows("Subgroup", 3, 5)
```
There are conflicts between the styling options
from `kableExtra` and the ones specified in `table_options()$CSS`;
some modifications might be needed to make everything work.  For
instance, the code above requests striping, but that
did not show up.  Experimentation may be needed!