---
title: "HTML Tables"
author: "Duncan Murdoch"
date: "January 17, 2016"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{HTML Tables}
  \usepackage[utf8]{inputenc}
---

This short R Markdown document illustrates how
to use `tables` in HTML. 

## Initializing

First, you need to load `tables`.  Packages
that it uses print banner messages, so you'll 
usually want to set a chunk option of `message=FALSE`.
```{r results='hide', message=FALSE}
library(tables)
```

The table is now constructed in the usual way.  
```{r}
X <- rnorm(125, sd=100)
Group <- factor(sample(letters[1:5], 125, rep=TRUE))
tab <- tabular( Group ~ (N=1)+Format(digits=2)*X*((Mean=mean) + Heading("Std Dev")*sd) )
```
In an R Markdown document, you don't want each table
to output the HTML document header, so turn 
off those options:
```{r}
table_options(htmloptions(head=FALSE))
```

## Inserting a Table

To format a table in HTML, use the `results = 'asis'` knitr option.  Before the first table in a document, also add a call to `writeCSS()` with `results = 'asis'`:
```{r results='asis'}
writeCSS()
html(tab)
```

The default justification makes the columns of 
numbers look messy.  You can set the justification
to the right, but the headers look wrong:
```{r results='asis'}
table_options(htmloptions(head = FALSE, justification = "r"))
html(tab)
```

The best look comes with the `pad = TRUE` option. 
This adds nonbreaking spaces around the numbers so that 
centering looks good.  It also changes the hyphens
to proper minus signs:
```{r results='asis'}
table_options(htmloptions(head = FALSE, justification = "c", pad = TRUE))
html(tab)
```
Unfortunately, if you cut this table and paste it
into a spreadsheet, the spaces and minus signs
probably won't be understood.  I don't know 
how to get everything we want :-(.

## Fine tuning

This document uses the default CSS from `table_options()$CSS`.  If you are
producing an `html_document`, it should be okay.
It does not look quite right in a
`slidy_presentation`, and is no good at all in
an `ioslides_presentation`.  Furthermore, you 
might not agree with my design choices.

In any of these cases, you should substitute your
own CSS.  You will need to modify the default 
one, and can use it as the `CSS` argument to 
`writeCSS()`, or set it as a new default in
`table_options()`.

Here is the default setting:
```{r comment=NA, echo=FALSE}
cat(table_options()$CSS)
```
Note that the `#ID` values will be replaced with
the `id` string given in `writeCSS()`.